A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-04-17 paper/paper1/ms.tex
Run text through grammarly

Uncertainty in system ages, evaporation timescales, and dominating physical mechanisms, degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of astrophysical variance could also subdue the appearance of atmospheric escape, even where we expect it most.  The 48 published non-detections of atmospheric escape \citep{2022arXiv221116243D} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources that seem predisposed to loss.  These intrinsic or extrinsic factors may include close proximity to host star, low surface gravity, and high energy incident radiation.  The small planet radius gap between 1.5$-$2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of eroding to their final rocky state.

@@

But inflated hot Saturns are rare \citep{2018AJ....155..214T}.  The cause for their underabundance remains an open question, with two leading candidate explanations.  Tidal migration mechanisms may be mass-dependent in such a way as to proceed efficiently for Jupiter-mass planets, but inefficiently for the lower mass Saturns.  In this scenario, sub-Saturn mass planets never make it to the close-in orbital separations that would lead to the conditions needed for inflation in the first place.

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may effectively migrate to close-in orbital separations, but once they arrive, the intense irradation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  \citet{2018AJ....155..214T} recognized the above two explanations for the lack of sub-Saturns, while statistically favoring Ohmic dissipation as the mechanism responsible for inflating Hot Jupiters.  Therefore Ohmic dissipation stands out as a leading physical mechanism unifying several observations.  Ohmic dissipation does not preclude the existence and role of other heating mechanisms---such as photoevaporation and tidal heating---but it does stand out has having made successful predictions before data were widely available to test them.

A runaway inflation scenario implies an inevitable and profound mass loss rate for inflated hot Saturns.  The planet's atmosphere overflows its Roche lobe at an ever increasing pace, yielding instantaneous mass loss rates possibly much larger than we have seen in previous long-term stable systems.  The ability to detect that conceivable mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own limitations.  Ly $\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity linecore, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages over Ly$\alpha$.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong instellation.  \hatpb---among the most inflated planets known---is an outlier.}
    \label{fig:instellation}
\end{figure}

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8 day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  It orbits a possibly evolved subgiant, opening the prospect for ``re-inflation'' \citep{2022AJ....163...53S,2022AJ....163..120G,2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase.  The tidal gravity of the nearby massive host star may accelerate mass loss rates even further \citep{2007A&A...472..329E}. Its status as a rare inflated Hot Saturn make \object{HAT-P-67 b} a promising laboratory, including but not limited to, testing the inflationary predictions of the Ohmic dissipation mechanism.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on the nights beginning 2020 April 28, 2020 May 22, 2020 June 15, and 2022 April 28.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transits were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit nights had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase low priority ``P4'' queue-filler observations.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total 41 visits on 39 unique nights, with two of those nights observing both an east and a west HET track.  The total on-source integration time exceeds 13.8 hours.  Table \ref{tabHPFLog} summarizes the log of observations.

The observation scheduling was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833$\AA~$ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium linewing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  Its coarse \emph{Gaia}-based estimates for RV and Doppler broadening are consistent with the more precise values published in \citet{2017AJ....153..211Z}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

\subsection{ASAS-SN, DASCH, and ZTF}
We retrieved ground-based photometry with the \emph{All-Sky Automated Survey for Supernovae} (ASAS-SN) using the Sky-Portal \citep{shappee14,2017PASP..129j4502K}.  The precision of ASAS-SN was too low to perceive stellar variability, and so we can place a relatively uninformative limit of $<5\%$ stellar variability on years-long timescales.

Similarly, \hatp appears in the Harvard Plate Archive, with 5809 measurements digitized through the Digital Access to a Sky Century at Harvard (DASCH) program, spanning over 120 years of coarse photometric monitoring.  In principle these datasets could inform long-term variability trends such as stellar cycles.  In practice the 0.15 magnitude jitter appears too coarse to perceive any genuine astrophysical variability, with no conspicuous trend seen.  We can therefore place a relatively mild constraint that the star is stable at the $\sim30\%$ level over periods of tens to hundreds of years.

@@

\subsection{Other archives}
\hatp has one $R\sim8000$ spectrum from the Intermediate Dispersion Spectrograph on the INT, centered near 4000 \AA.  It has 26 photometry points from \emph{Pan-STARRS} spread across its 5 filters.  The other archival observations have been previously reported by \citet{2017AJ....153..211Z}, including Keck HIRES, Keck NIRC2, and WIYN High-Resolution Infrared Camera.  Overall the system appears relative unstudied outside of the discovery paper.

@@

\subsection{Gaia DR3}\label{gaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere 0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint, and from the model dependence of multiple orbit fits.

\subsection{TESS Light Curve}
Previously, \hatpb~transits had only been detected with \emph{HATNet} \citep{2004PASP..116..266B} and followed up with KeplerCam on the FLWO 1.2 m telescope \citep{2017AJ....153..211Z}. These ground-based photometers were not intended to measure long term stellar variability signals.  We therefore examined the \emph{TESS} lightcurves for out-of-transit photometric variability.  The revised precision and continuous coverage of \emph{TESS} can refine the orbital solution.

\subsubsection{Revised exoplanet orbital parameters}
We assembled a composite lightcurve by stitching TESS Sectors 24, 26, 51, 52, 53, which were reduced with the default SPOC pipeline \citep{2020RNAAS...4..201C}, and lightly post-processed with \texttt{lightkurve} \citep{geert_barentsen_2019_2565212}.  These TESS data exhibit an RMS scatter of better than 1 part-per-thousand (ppt) at native 2-minute sampling.  We fit a Keplerian orbit model to the TESS lightcurve using the \texttt{exoplanet} framework \citep{exoplanet:joss}.  We obtained a best fit orbital period of $4.81011$ days, non-zero eccentricity of 0.19, an impact parameter of 0.46, planet-to-star radius ratio of 0.0823, transit depth of 0.74\%, and $T_0$ of 1958.07933 in BTJD.  These properties are consistent with the previously reported values from \citet{2017AJ....153..211Z} and the updated ephemeris of \citet{2022ApJS..259...62I}.  Figure \ref{fig:TESSoverview} shows an overview of all the TESS Sectors demarcated with the epochs of HPF measurements.  Figure \ref{fig:transit} shows the best-fit orbit overlaid on the detrended TESS lightcurve.  Table \ref{tabOrbit} lists one of the previous orbit determinations and solution reported here.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation on timescales comparable to the exoplanet orbit. TESS Sectors 51-53 do not show as conspicuous a modulation signal.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a period of 4.7 days; when fit to individual sectors alone the periods hover around 5.9 days.

If we attribute this TESS modulation signal to stellar activity, we obtain a rotation period $P_\mathrm{rot}$ between 4.7 and 5.9 days.  We can constrain the expected rotation period based on our revised stellar radius estimate, stellar rotation rate from $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range may be typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7 day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9 day per-Sector fits, slightly preferring the lower 4.7 day modulation as the stellar rotation period.  The 5.9 day value would require a stellar radius over 3.2$R_\odot$, which appears consistent with only an eccentric orbit model combined with the Geneva isochrone \citep{2017AJ....153..211Z}.

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
As discussed in Section \ref{gaiaDR3}, the slightly increased distance and radius implies a slightly more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;]{2016ApJS..222....8D,2016ApJ...823..102C} for \hatp.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation to \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.
\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{figures/MIST_tracks_HATP.png}
    \caption{MIST evolutionary model tracks for \hatp.  The $T_\mathrm{eff}$ and luminosity of \hatp~ (gray circle) are consistent with either the late main sequence or recently evolved subgiant.}
    \label{fig:evolTracks}
\end{figure*}

@@

We reduced and analyzed the HPF data first for the purpose of orbit fitting via precision radial velocity measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves, and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$, slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  Notably our fit does not jointly model the Doppler tomographic signals reported in \citep{2017AJ....153..211Z}, which would further constrain the impact parameter.  Additionally our input $\rstar$ was slightly lower than the best fit value derived from \emph{Gaia DR3} and SED fitting.
\authorcomment1{Quang Tran: please verify and revise.}

@@

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV, namely we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D echellogram reduction.  This tool outputs 1D extracted spectra for the target, and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were usually acquired with the LFC turned off, so this unilluminated spectrum was discarded.

>>>>

Uncertainty in system ages, evaporation timescales, and dominating physical mechanisms, degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of astrophysical variance could also subdue the appearance of atmospheric escape, even where we expect it most.  The 48 published non-detections of atmospheric escape \citep{2022arXiv221116243D} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources that seem predisposed to loss.  These intrinsic or extrinsic factors may include proximity to the host star, low surface gravity, and high energy incident radiation.  The small planet radius gap between 1.5$-$2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of eroding to their final rocky state.

@@

But inflated hot Saturns are rare \citep{2018AJ....155..214T}.  The cause for their underabundance remains an open question, with two leading candidate explanations.  Tidal migration mechanisms may be mass-dependent in such a way as to proceed efficiently for Jupiter-mass planets, but inefficiently for the lower-mass Saturns.  In this scenario, sub-Saturn mass planets never make it to the close-in orbital separations that would lead to the conditions needed for inflation in the first place.

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may effectively migrate to close-in orbital separations, but once they arrive, the intense irradiation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  \citet{2018AJ....155..214T} recognized the above two explanations for the lack of sub-Saturns, while statistically favoring Ohmic dissipation as the mechanism responsible for inflating Hot Jupiters.  Therefore Ohmic dissipation stands out as a leading physical mechanism unifying several observations.  Ohmic dissipation does not preclude the existence and role of other heating mechanisms---such as photoevaporation and tidal heating---but it does stand out as having made successful predictions before data were widely available to test them.

A runaway inflation scenario implies an inevitable and profound mass loss rate for inflated hot Saturns.  The planet's atmosphere overflows its Roche lobe at an ever-increasing pace, yielding instantaneous mass loss rates possibly much larger than we have seen in previous long-term stable systems.  The ability to detect that conceivable mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own limitations.  Ly $\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages over Ly$\alpha$.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong insolation.  \hatpb---among the most inflated planets known---is an outlier.}
    \label{fig:instellation}
\end{figure}

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  It orbits a possibly evolved subgiant, opening the prospect for ``re-inflation'' \citep{2022AJ....163...53S,2022AJ....163..120G,2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase.  The tidal gravity of the nearby massive host star may accelerate mass loss rates even further \citep{2007A&A...472..329E}. Its status as a rare inflated Hot Saturn makes \object{HAT-P-67 b} a promising laboratory, including but not limited to, testing the inflationary predictions of the Ohmic dissipation mechanism.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on the nights beginning 2020 April 28, 2020; May 22, 2020; June 15, 2020; and 2022 April 28.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transits were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit nights had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase low-priority ``P4'' queue-filler observations.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total of 41 visits on 39 unique nights, with two of those nights observing both an east and a west HET track.  The total on-source integration time exceeds 13.8 hours.  Table \ref{tabHPFLog} summarizes the log of observations.

The observation schedule was strategically restricted to the spring season when the Barycentric Earth Radial Velocity (BERV) would Doppler shift telluric absorption lines sufficiently far away from the core of the Helium 10833$\AA~$ feature \citep{2022ApJ...939L..11S}.  The small BERV still means the redward Helium line wing has significant telluric contamination between 10834$-$10836 \AA, but the core and blue line wings appear relatively pristine.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  Its coarse \emph{Gaia}-based estimates for RV and Doppler broadening are consistent with the more precise values published in \citet{2017AJ....153..211Z}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

\subsection{ASAS-SN, DASCH, and ZTF}
We retrieved ground-based photometry with the \emph{All-Sky Automated Survey for Supernovae} (ASAS-SN) using the Sky-Portal \citep{shappee14,2017PASP..129j4502K}.  The precision of ASAS-SN was too low to perceive stellar variability, so we can place a relatively uninformative limit of $<5\%$ stellar variability on years-long timescales.

Similarly, \hatp appears in the Harvard Plate Archive, with 5809 measurements digitized through the Digital Access to a Sky Century at Harvard (DASCH) program, spanning over 120 years of coarse photometric monitoring.  In principle, these datasets could inform long-term variability trends such as stellar cycles.  In practice, the 0.15 magnitude jitter appears too coarse to perceive any genuine astrophysical variability, with no conspicuous trend seen.  We can therefore place a relatively mild constraint that the star is stable at the $\sim30\%$ level over periods of tens to hundreds of years.

@@

\subsection{Other archives}
\hatp has one $R\sim8000$ spectrum from the Intermediate Dispersion Spectrograph on the INT, centered near 4000 \AA.  It has 26 photometry points from \emph{Pan-STARRS} spread across its 5 filters.  The other archival observations have been previously reported by \citet{2017AJ....153..211Z}, including Keck HIRES, Keck NIRC2, and WIYN High-Resolution Infrared Camera.  Overall the system appears relatively unstudied outside of the discovery paper.

@@

\subsection{Gaia DR3}\label{gaiaDR3}
\citet{2017AJ....153..211Z} previously derived stellar radius estimates of 2.1-2.7 $R_\odot$ through Spectral Energy Distribution (SED) isochrone fitting as part of a joint orbit fit.  We systematically increase those stellar radius estimates by 8.7\% to match the greater Gaia \emph{DR3} distance (\S \ref{gaiadr3}).  For a fixed $R_p/R_\star$, the larger $R_\star$ implies a proportionally larger planet radius. This update systematically decreases the estimate for the already-low density of HAT-P-67 by 28\%, to a mere 0.035 g$\;$cm$^{-3}$, albeit with significant uncertainties from the weak mass constraint.

\subsection{TESS Light Curve}
Previously, \hatpb~transits had only been detected with \emph{HATNet} \citep{2004PASP..116..266B} and followed up with KeplerCam on the FLWO 1.2 m telescope \citep{2017AJ....153..211Z}. These ground-based photometers were not intended to measure long-term stellar variability signals.  We, therefore, examined the \emph{TESS} lightcurves for out-of-transit photometric variability.  The revised precision and continuous coverage of \emph{TESS} can refine the orbital solution.

\subsubsection{Revised exoplanet orbital parameters}
We assembled a composite lightcurve by stitching TESS Sectors 24, 26, 51, 52, 53, which were reduced with the default SPOC pipeline \citep{2020RNAAS...4..201C}, and lightly post-processed with \texttt{lightkurve} \citep{geert_barentsen_2019_2565212}.  These TESS data exhibit an RMS scatter of better than 1 part-per-thousand (ppt) at native 2-minute sampling.  We fit a Keplerian orbit model to the TESS lightcurve using the \texttt{exoplanet} framework \citep{exoplanet:joss}.  We obtained a best fit orbital period of $4.81011$ days, non-zero eccentricity of 0.19, an impact parameter of 0.46, planet-to-star radius ratio of 0.0823, transit depth of 0.74\%, and $T_0$ of 1958.07933 in BTJD.  These properties are consistent with the previously reported values from \citet{2017AJ....153..211Z} and the updated ephemeris of \citet{2022ApJS..259...62I}.  Figure \ref{fig:TESSoverview} shows an overview of all the TESS Sectors demarcated with the epochs of HPF measurements.  Figure \ref{fig:transit} shows the best-fit orbit overlaid on the detrended TESS lightcurve.  Table \ref{tabOrbit} lists one of the previous orbit determinations and the orbital solution reported here.

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation on timescales comparable to the exoplanet orbit. TESS Sectors 51-53 do not show as conspicuous a modulation signal.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.

If we attribute this TESS modulation signal to stellar activity, we obtain a rotation period $P_\mathrm{rot}$ between 4.7 and 5.9 days.  We can constrain the expected rotation period based on our revised stellar radius estimate, stellar rotation rate from $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range may be typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears consistent with only an eccentric orbit model combined with the Geneva isochrone \citep{2017AJ....153..211Z}.

@@

\subsection{Revised system evolutionary state}\label{secMISTtracks}
As discussed in Section \ref{gaiaDR3}, the slightly increased distance and radius imply a slightly more luminous host star than previously estimated.  Figure \ref{fig:evolTracks} shows evolutionary tracks using the solar-metallicity MESA \citep{2011ApJS..192....3P,2013ApJS..208....4P,2015ApJS..220...15P} Isochrones and Stellar Tracks \citep[MIST;]{2016ApJS..222....8D,2016ApJ...823..102C} for \hatp.  The tracks show that a 1.8 Gyr, 1.64 $M_\odot$ \hatp~ could be at the end of its main sequence lifetime (teal track), yielding a gradual rise in incident radiation to \hatpb.  Alternatively, a 1.54 $M_\odot$ evolutionary track would send \hatp~ along the subgiant branch at an age over 2.2 Gyr, with a rapid 50\% increase in luminosity, potentially leading to ``re-inflation'' of the planet \citep{2021ApJ...909L..16T}.  Either evolutionary track appears consistent with the observed SED, stellar surface gravity, and available constraints from spectral fitting.
\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{figures/MIST_tracks_HATP.png}
    \caption{MIST evolutionary model tracks for \hatp.  The $T_\mathrm{eff}$ and luminosity of \hatp~ (gray circle) are consistent with either the late main sequence or recently evolved subgiant.}
    \label{fig:evolTracks}
\end{figure*}

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$, slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.  Notably, our fit does not jointly model the Doppler tomographic signals reported in \citep{2017AJ....153..211Z}, which would further constrain the impact parameter.  Additionally, our input $\rstar$ was slightly lower than the best-fit value derived from \emph{Gaia DR3} and SED fitting.
\authorcomment1{Quang Tran: please verify and revise.}

@@

\subsection{HPF analysis II: \ion{He}{1} 10833 \AA}
The preparation for analyzing the Helium feature followed a similar but slightly different procedure than that used for the RV, namely, we used \texttt{Goldilocks}\footnote{\url{https://github.com/grzeimann/Goldilocks_Documentation}} for 2D echellogram reduction.  This tool outputs 1D extracted spectra for the target, and two reference fibers: blank sky and a laser frequency comb \citep[LFC,][]{2019Optic...6..233M}.  The observations were usually acquired with the LFC turned off, so this unilluminated spectrum was discarded.
